---
title: "Working w dates and regex"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = "/Users/rebecca/Desktop/DS19-class/ds19-class/")

install.packages("lubridate")
install.packages("dplyr")
install.packages("ggplot2")
library(lubridate)
library(dplyr)
library(ggplot2)
```

## Purpose
Explore BikeTown data looking at mean miles traveled per week during the 2018 summer.

``` {r pull_data}
source("/Users/rebecca/Desktop/DS19-class/ds19-class/code/fetch_biketown.R")
get_data(start = "06/2018", end = "09/2018", outdir = "/Users/rebecca/Desktop/DS19-class/ds19-class/data/biketown/")
# needed to set the outdir

# to pull in all of the files at once with the source
# set working drive to actual location where the data is sourced

setwd("/Users/rebecca/Desktop/DS19-class/ds19-class/data/biketown/")
folder <- "/Users/rebecca/Desktop/DS19-class/ds19-class/data/biketown/"
filenames <- list.files(path = folder, pattern = "*.csv", all.files = FALSE, 
                        full.names = FALSE, recursive = FALSE, ignore.case = FALSE)
read_csv_filename <- function(filenames) {
  ret <- read.csv(filenames, stringsAsFactors = F, strip.white = T, na.strings = "")
  ret$Source <- filenames
  ret
}
bike_raw <- plyr::ldply(filenames, read_csv_filename)
```

```{r data_wrangling}
str(bike_raw)
# str stands for structure

bike_df1 <- bike_raw %>% 
  mutate(start.datetime = paste(StartDate, StartTime, sep = " "),
        end.datetime = paste(EndDate, EndTime, sep = " "))
# paste0 means no spaces in between

str(bike_df1)

# convert start.datetime and end.datetime into date-time format,
# make sure to include timezone! 
# check acceptable timezone values
OlsonNames()
# 587, US Pacific

# checking the time/date formatting
head(bike_df1$start.datetime)

# hm for hour:minutes, hms for hour: minute:sec
bike_df1$start.datetime <- mdy_hm(bike_df1$start.datetime, tz = "America/Los_Angeles")
bike_df1$end.datetime <- mdy_hm(bike_df1$end.datetime, tz = "America/Los_Angeles")
str(bike_df1)

# convert duration into a useable format
bike_df1$Duration <- hms(bike_df1$Duration)
# Error thrown after converting to duration indicating NAs in the duration column,
# Check raw dataframe for NAs in the Duration column
sum(is.na(bike_raw$Duration))
# important to check raw data, as the hms function in bike_df1 rounded everything to 1 minute.
```
`lubridate` has three synonomous functions, but they define very different actions:

1. duration(): time spans are in units of seconds and there is no start date involved.
2. interval(): also measured in units of seconds, provides the difference between start and end times.
3. period(): measures time span in units larger than seconds. This is handy when accounting for Daylight Savings times and Leap Years.

``` {r duration_calc}
# Caltulate interval
bike_df1$interval <- interval(bike_df1$start.datetime, bike_df1$end.datetime)
head(bike_df1$interval)

# Calulate duration - retains from the interval function all aspects of time and converts to seconds
bike_df1$duration.all <- as.duration(bike_df1$interval)
head(bike_df1$duration.all)

# could rewrite this as a nested or piped function

# Calculate period - provides answers in days, hours, minutes, seconds - higher resolution
bike_df1$period <- as.period(bike_df1$duration.all)
head(bike_df1$period)

# lubridate also handles time aggregation really well
# floor date() is helpful for aggregating 
bike_wkagg <- bike_df1 %>% 
  mutate(week.datetime = floor_date(start.datetime, unit = "week"))
# can now group by the date
head(bike_wkagg$week.datetime)

# also, ceiling date, rolling date
mean_dist_wk <- bike_wkagg %>% 
  group_by(week.datetime) %>% 
  summarise(mean_miles = mean(Distance_Miles))
mean_dist_wk

# one way, simpler
mdw_fig <- mean_dist_wk %>% 
  ggplot(aes(x = week.datetime, y = mean_miles, fill = week.datetime)) + 
  geom_bar(stat = "identity")

# second way, a little more complex
mdw_fig2 <- mean_dist_wk %>% 
  ggplot(aes(x = as.Date(week.datetime), y = mean_miles)) + 
  geom_bar(stat = "identity", fill = "blue") + 
  scale_x_date(date_breaks = "1 week") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
mdw_fig2

# ggplot will assume continuous; if we are missing data, 
# we will need to insert a break to signify where the data is missing
# can add a baseline, add a block/rectangle to emphasize results

```




